---
title: "COVID-19 Data Analysis"
author: "Primal Patel"
date: "2026-08-12"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction and Questions of Interest

This report analyzes the Johns Hopkins University CSSE COVID-19 dataset, which
tracks confirmed cases and deaths from January 22, 2020 through March 9, 2023,
when JHU stopped collecting data. The data comes as four time series files in
wide format: one row per location (country or province globally; county for the
US) and one column per date, holding *cumulative* counts. A separate lookup
table supplies populations, which we join in to enable per-capita comparisons.

The analysis addresses three questions:

1. How did cases and deaths evolve over time in the US, and what distinct waves
   are visible?
2. How does the US compare to other countries, and how much do outcomes vary
   across US states, on a per-capita basis?
3. Does a state's case rate predict its death rate? ie. If we know
   how many infections a state had per capita, how well can we predict its
   deaths per capita?

**Required libraries:** `tidyverse` and `lubridate`.

```{r libraries}
library(tidyverse)
library(lubridate)
```

## Importing the Data

The data comes from the Johns Hopkins CSSE COVID-19 repository on GitHub. We
read in four time series files (global and US, cases and deaths) plus a lookup
table containing population data. Reading directly from the URLs keeps the
analysis fully reproducible.

```{r import_data}
url_in <- paste0("https://raw.githubusercontent.com/CSSEGISandData/",
                 "COVID-19/master/csse_covid_19_data/",
                 "csse_covid_19_time_series/")

file_names <- c("time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_global.csv",
                "time_series_covid19_confirmed_US.csv",
                "time_series_covid19_deaths_US.csv")

urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[2])
us_cases <- read_csv(urls[3])
us_deaths <- read_csv(urls[4])

uid_lookup_url <- paste0("https://raw.githubusercontent.com/CSSEGISandData/",
                         "COVID-19/master/csse_covid_19_data/",
                         "UID_ISO_FIPS_LookUp_Table.csv")
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

## Tidying and Transforming the Data

The raw files are in wide format, with one column per date. We pivot each date
column into rows, convert the dates from character to date objects, drop the
latitude/longitude columns we will not use, and join cases with deaths. For the
global data we also join in populations from the lookup table so per-capita
comparisons are possible.

```{r tidy_global}
global_cases <- global_cases %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "cases") %>%
  select(-c(Lat, Long))

global_deaths <- global_deaths %>%
  pivot_longer(cols = -c(`Province/State`, `Country/Region`, Lat, Long),
               names_to = "date",
               values_to = "deaths") %>%
  select(-c(Lat, Long))

global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date)) %>%
  filter(cases > 0) %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE) %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, Population, Combined_Key)

summary(global)
```

```{r tidy_us}
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Population),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us <- us_cases %>%
  full_join(us_deaths)

summary(us)
```

Next we aggregate the US county-level data up to the state level and to the
national level, and calculate deaths per million for per-capita comparison.

```{r us_by_state}
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

us_totals <- us_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths * 1000000 / Population) %>%
  select(Country_Region, date,
         cases, deaths, deaths_per_mill, Population) %>%
  ungroup()

tail(us_totals)
```

By the end of the dataset (March 9, 2023), the US had recorded roughly 103.8
million cases and 1.12 million deaths, out of a global total of about 676.6
million cases and 6.88 million deaths.

## Visualizing the Data

### Cumulative cases and deaths in the US

```{r visualize_us_totals}
us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID-19 in the US: Cumulative Cases and Deaths",
       y = "count (log scale)", x = NULL, color = NULL)
```

On the log scale, both curves rise rapidly in US Spring 2020 and then flatten.
Cumulative counts, however, hide the wave structure of the pandemic, so we look
at *new* cases instead. The coverage of why the wave-structure of the pandemic is important has an example here: <https://www.cell.com/heliyon/fulltext/S2405-8440(24)01121-6?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS2405844024011216%3Fshowall%3Dtrue>. The wave structure indicates both the evolution and expected progression of the virus. 

### New cases: the wave structure of the pandemic

Because cases and deaths are cumulative, we compute daily new counts with
`lag()`, then smooth with a 7-day rolling average to remove the strong
day-of-week reporting cycle (a bias listed below).

```{r visualize_new_cases}
us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

us_totals <- us_totals %>%
  mutate(new_cases_7day = stats::filter(new_cases, rep(1/7, 7), sides = 1),
         new_deaths_7day = stats::filter(new_deaths, rep(1/7, 7), sides = 1))

us_totals %>%
  filter(!is.na(new_cases_7day)) %>%
  ggplot(aes(x = date, y = as.numeric(new_cases_7day))) +
  geom_line(color = "steelblue") +
  theme(legend.position = "bottom") +
  labs(title = "US Daily New Cases (7-day average)",
       y = "new cases per day", x = NULL)

us_totals %>%
  filter(!is.na(new_deaths_7day)) %>%
  ggplot(aes(x = date, y = as.numeric(new_deaths_7day))) +
  geom_line(color = "firebrick") +
  labs(title = "US Daily New Deaths (7-day average)",
       y = "new deaths per day", x = NULL)
```

Several distinct waves are visible: spring 2020, summer 2020, the large winter
2020-21 wave, the Delta wave in late summer 2021, and the enormous Omicron
spike, which peaked at roughly 807,000 new cases per day in mid-January 2022.
Notably, the deaths curve does **not** mirror the cases curve: daily deaths
peaked at about 3,400 per day in January 2021, a year *before* the case peak.
The Omicron wave produced far more cases but proportionally fewer deaths,
consistent with vaccination, prior immunity, and a less lethal variant.

### Case fatality ratio over time

We can separate the wave-structure and fatality by looking at the cumulative case fatality ratio (CFR),
deaths divided by cases. See: <https://www.britannica.com/science/case-fatality-rate>

```{r cfr}
us_totals %>%
  filter(cases > 1000) %>%
  mutate(cfr = deaths / cases) %>%
  ggplot(aes(x = date, y = cfr)) +
  geom_line(color = "darkred") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "US Cumulative Case Fatality Ratio",
       y = "deaths / cases", x = NULL)
```

The cumulative CFR fell from about 1.7% at the end of 2020 to about 1.1% by the
end of the dataset. The early figure is inflated by limited testing (many mild
cases were never counted), and the later decline reflects vaccines, better
treatment, and the milder Omicron variant.

### How does the US compare globally?

Using the tidied global data (with populations joined in), we compute each
country's cumulative deaths per thousand residents and compare the US against a
set of peer countries, restricting to countries with over one million people so
small-population outliers don't dominate.

```{r global_comparison}
country_totals <- global %>%
  group_by(Country_Region) %>%
  filter(date == max(date)) %>%
  summarize(cases = sum(cases), deaths = sum(deaths),
            Population = sum(Population, na.rm = TRUE)) %>%
  filter(Population > 1000000) %>%
  mutate(deaths_per_thou = 1000 * deaths / Population,
         cases_per_thou = 1000 * cases / Population)

comparison_countries <- c("US", "United Kingdom", "Italy", "France",
                          "Germany", "Canada", "Japan", "Korea, South",
                          "Mexico", "Brazil", "India", "Peru",
                          "Australia", "Sweden")

country_totals %>%
  filter(Country_Region %in% comparison_countries) %>%
  ggplot(aes(x = reorder(Country_Region, deaths_per_thou),
             y = deaths_per_thou,
             fill = Country_Region == "US")) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = c("grey60", "firebrick")) +
  coord_flip() +
  labs(title = "COVID-19 Deaths per Thousand: US vs. Peer Countries",
       x = NULL, y = "deaths per thousand residents")

country_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(Country_Region, deaths_per_thou, cases_per_thou)
```

At about 3.4 deaths per thousand, the US fared worse than nearly every wealthy
peer — the UK, Italy, and France sit near or below 3.2, Germany at 2.0, Canada
at 1.4, and Japan and South Korea below 0.7 — and ranks roughly 12th worst
among the 157 countries with over a million people, in the company of Peru
(6.7, the world's worst) and the hardest-hit countries of Eastern Europe. This
comparison must be read cautiously: reported deaths dramatically undercount
true mortality in countries with weaker reporting fidelity, so the honest conclusion is
that the US did poorly *relative to countries with comparable reporting
systems*, not that it did worse than most of the world.

### Which states fared best and worst?

```{r state_extremes}
us_state_totals <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000 * cases / population,
            deaths_per_thou = 1000 * deaths / population) %>%
  filter(cases > 0, population > 0)

us_state_totals %>%
  slice_max(deaths_per_thou, n = 10) %>%
  select(Province_State, deaths_per_thou, cases_per_thou)

us_state_totals %>%
  slice_min(deaths_per_thou, n = 10) %>%
  select(Province_State, deaths_per_thou, cases_per_thou)
```

Death rates vary by roughly a factor of three among the 50 states: Arizona,
Oklahoma, and Mississippi all exceeded 4.4 deaths per thousand residents, while
Hawaii and Vermont stayed below 1.5. The national
average was about 3.4 deaths per thousand.

## Modeling the Data

Does knowing a state's case rate let us predict its death rate? We fit a simple
linear model of deaths per thousand as a function of cases per thousand across
the states and territories.

```{r model}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = us_state_totals)
summary(mod)

us_state_totals <- us_state_totals %>%
  mutate(pred = predict(mod))

us_state_totals %>%
  ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou),
             color = "blue") +
  geom_line(aes(x = cases_per_thou, y = pred), color = "red") +
  labs(title = "State Deaths per Thousand vs. Cases per Thousand",
       subtitle = "Blue = actual, red = linear model prediction",
       x = "cases per thousand", y = "deaths per thousand")
```

The relationship is positive and statistically significant, but weak: the model
explains only about 31% of the variance (R-squared of roughly 0.31). Case rates
alone are a poor predictor of death rates. States far above the line (more
deaths than their case count predicts) and far below it (fewer) point to other
drivers worth investigating: age structure, vaccination rates, hospital
capacity, testing intensity, and the timing of when each state was hit — states hit hard in spring
2020 faced the virus before treatments and vaccines existed.

```{r model_residuals}
us_state_totals %>%
  mutate(resid = deaths_per_thou - pred) %>%
  slice_max(abs(resid), n = 6) %>%
  select(Province_State, cases_per_thou, deaths_per_thou, pred, resid)
```

## Potential Sources of Bias

Several sources of bias should temper the conclusions:

**Measurement bias in case counts.** Confirmed cases depend on testing.
Testing was scarce in early 2020 and changed toward the use of home antigen tests
(rarely reported) in 2022, so case undercounting varies over time and by state.
This directly affects both the CFR analysis and the regression: some of the
state-to-state variation in "cases per thousand" is variation in testing, not
in infection.

**Measurement bias in death counts.** States differed in how they attributed
deaths to COVID-19 (deaths *with* versus *from* COVID), and death
certificate practices changed over the pandemic.

**Reporting artifacts.** The raw data contain day-of-week cycles and
occasional bulk corrections (including negative "new case" days when
jurisdictions revised totals). The 7-day average mitigates but does not
eliminate this.

**International comparability.** Countries differ enormously in testing
capacity, death registration systems, and political incentives to report
accurately. Cross-country comparisons of *reported* rates are only meaningful
among countries with similar surveillance quality, which is why the global
comparison above is framed around wealthy peer countries.

**State model** The state-level regression cannot be interpreted at
the individual level; it also treats each state as one point, ignoring huge
within-state variation between counties.

**Personal bias.** I came to this analysis expecting vaccination-era waves to
be less deadly, which is a conclusion the data supports, but I have tried to
mitigate confirmation bias by reporting the model's weak fit honestly and by
listing alternative explanations (testing intensity, age structure, timing)
rather than attributing state differences to any single cause.

## Conclusion

The JHU data show a pandemic that arrived in waves, with the link between
infections and deaths weakening dramatically over time: the January 2022
Omicron wave produced the most cases by far, yet daily deaths peaked a full
year earlier, and the cumulative CFR fell from about 1.7% to 1.1%. On a
per-capita basis the US recorded more deaths than nearly all of its wealthy
peers, though cross-country comparisons are limited by reporting differences.
Outcomes
across states varied threefold on a per-capita basis, and a state's case rate
explains only about a third of the variance in its death rate — meaning most of
what determined how deadly the pandemic was in a given state lies in factors
beyond raw infection counts. A natural follow-up would be to join in state-level
vaccination, age, and hospital-capacity data to model those residuals directly.

I acknowledge the use of AI (Claude Code) to help me check my syntactic constructions in building the code, to search for the correct voacbulary to implementing my intentions where I lacked the choice of right libraries, programming strategy, the need to write a function over the larger dataset, and to help me find the right verb when I needed it. I also used a search engine to help generate ideas for questions to be posed regarding policing work. I relied on the visualisation videos from module 1 to give inspiration to the questions and analysis. 

Sample:
Thought process
Thought process
All good questions — let me audit the document against each. The honest answer: six of seven are solidly covered, but the global-vs-US comparison is the weak spot. We import and tidy the global data but only cite world totals in prose — we never actually compare rates. Let me verify numbers for a proper comparison and fix that.

"
]
Rubric audit, item by item:

Global vs US rates — this was the one gap. We imported global data but only cited totals in prose. Now fixed: a new "How does the US compare globally?" section computes deaths per thousand by country, plots the US against 13 peer countries, and tables the 10 worst (population > 1M). Verified numbers: US ~3.4/thousand, roughly 12th worst of 157 countries — behind Peru (6.7) and Eastern Europe, far worse than Germany (2.0), Canada (1.4), Japan/Korea (<0.7).
Stated purpose — yes, "Introduction and Questions of Interest" poses three explicit questions (question 2 now includes the global comparison).
Data description — strengthened: the intro now specifies the source, date range, wide format (one row per location, one cumulative-count column per date, county-level for US), and the population lookup table.
Model — yes, lm(deaths_per_thou ~ cases_per_thou) across states with prediction plot, R² discussion, and residual table.
Personal bias — yes you have an explicit paragraph on prior expectation and mitigation.
Reporting/other biases — yes: testing-dependent case counts, death attribution differences, day-of-week/bulk-correction artifacts, ecological fallacy, and a new international-comparability paragraph (why we compare only against similar-surveillance peers — e.g., India's reported 0.4/thousand vs excess-mortality estimates).
sessionInfo() — yes, final chunk."

```{r session_info}
sessionInfo()
```
